---
title: "data_download_ranunculus"
author: "Hannah Pilat"
date: "2024-01-30"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a workflow for downloading occurrence data for *Ranunculus
glaberrimus* and environmental and anthropogenic predictor data (present
and future). These will be processed and used in a trial species
distribution model to get a realized and potential distribution/to model
suitable habitat for *R. glaberrimus*. This workflow will then be
repeated for *Claytonia lanceolata*, our focal species.

Start by creating directories for scripts, data, and outputs

```{r, echo = FALSE}
dir.create("data/")
dir.create("data/raw")
dir.create("data/processed")
dir.create("scripts/")
dir.create("outputs/")
```

All downloaded .tif and .shp files have been placed in the data/raw/ folder

Now upload required packages for data wrangling and processing:

```{r, echo = FALSE}
library(tidyverse)
library(geodata) # won't work when on AAFC VPN
library(terra)
library(raster)
library(sf)
library(rgbif)
```

### **Occurrence Data**

Download *Ranunculus glaberrimus* occurrence data from GBIF, using rgbif
package First, set up GBIF credentials/login information \^needs some
work to be reproducible

```{r, echo = FALSE}
install.packages("usethis")
usethis::edit_r_environ()
```

Download occurrence data for Ranunculus glaberrimus within North
America, between the years 1950-2024

```{r, echo = TRUE}
rgbif::occ_download(
  pred("hasGeospatialIssue", FALSE), 
  pred("hasCoordinate", TRUE),
  pred("continent", "north_america"),
  pred("year", "1940,2024"),
  pred("taxonKey", 3033299), 
  format = "SIMPLE_CSV")
```

Check the status of the occurrence download:

```{r, echo = TRUE}
occ_download_wait('0067361-231120084113126')
```

To access download when it's finished:

Note: I moved the CSV file to data/raw/ from the zipped GBIF folder but the below code isn't working

```{r, echo = TRUE}
ran_occ_download <- occ_download_get(key = '0067361-231120084113126', 
                                     path = "data/") %>%
  occ_download_import(key = '0067361-231120084113126', 
                      path = "data/")
```

Download Info: Username: hpilat E-mail:
[hannahepilat\@gmail.com](mailto:hannahepilat@gmail.com){.email} Format:
SIMPLE_CSV Download key: 0067361-231120084113126 Created:
2024-01-21T21:44:46.652+00:00 Citation
[[Info:\\\\](Info:){.uri}](%5BInfo:%5D(Info:)%7B.uri%7D){.uri} Please
always cite the download DOI when using this data.
<https://www.gbif.org/citation-guidelines> DOI: 10.15468/dl.3f98w4
Citation: GBIF Occurrence Download <https://doi.org/10.15468/dl.3f98w4>
Accessed from R via rgbif (<https://github.com/ropensci/rgbif>) on
2024-01-21

### **Predictor Data**

For our model, we want to input our predictor data as a multilayered
SpatRaster. Most of the data is available as GeoTiff (.tif) files, which
will be imported to R as SpatRasters. Some of the data come as
shapefiles, which will need to be vectorized and subsequently
rasterized.

#### **Present:**

Elevation data for North America (2023), downloaded from
<http://www.cec.org/north-american-environmental-atlas/elevation-2023/>

```{r, echo = FALSE}
elevation_na <- rast("data/raw/northamerica_elevation_cec_2023.tif")
```

Landcover data for North America (2020), downloaded from:
<http://www.cec.org/north-american-environmental-atlas/land-cover-30m-2020/>

```{r, echo = FALSE}
lndcvr_na <- rast("data/raw/NA_NALCMS_landcover_2020_30m.tif")
```

Global soil temperature seasonality data (standard deviation X 100), for
0-5 and 5-15 cm under the surface, which is relevant to where
*skwenkwinem* grows. Downloaded from:
<https://zenodo.org/records/7134169>

```{r, echo = FALSE}
soil_temp_0_5 <- rast("data/raw/SBIO4_0_5cm_Temperature_Seasonality.tif")
soil_temp_5_15 <- rast("data/raw/SBIO4_5_15cm_Temperature_Seasonality.tif")
```

Soil pH data for 0-5 and 5-15 cm below the surface, through the geodata
package. Depth = 5 indicates 0-5cm below surface Depth = 15 indicates
5-15cm below surface Downloads a folder called "soil_world"

```{r, echo = FALSE}
soil_phh2o_0_5 <- geodata::soil_world(var = "phh2o", depth = 5, stat =
"mean", path = "data/raw/")

soil_phh2o_5_15 <- geodata::soil_world(var = "phh2o", depth = 15, stat
= "mean", path = "data/raw/")
```

Global WorldClim monthly average temperature data at 2.5 arcsec
resolution, downloads a folder called "climate"

```{r, echo = FALSE}
temp_avg_global <- geodata::worldclim_global(var = "tavg", res = "2.5",
                                             path = "data/raw/", version = "2.1")
```

Global WorldClim monthly precipitation data at 2.5 arcsec resolution

```{r, echo = FALSE}
precip_global <- geodata::worldclim_global(var = "prec", res = "2.5",
                                           path = "data/raw/", version = "2.1")
```

Anthropogenic biome data (2008), downloaded from
<http://www.cec.org/north-american-environmental-atlas/anthropogenic-biomes/>

```{r, echo = FALSE}
anth_biome <- rast("data/raw/anthromes_EqArea.tif")
```

The following are available as shapefiles, and therefore we turn them
into SpatVectors in order to process them (later) to SpatVectors.

Watersheds data (2011), downloaded from
<http://www.cec.org/north-american-environmental-atlas/watersheds/>

```{r, echo = FALSE}
watersheds_sf <- read_sf("data/raw/watersheds_shapefile/Watersheds_Shapefile/NA_Watersheds/data/watershed_p_v2.shp")
watersheds_vect <- vect(watersheds_sf)
```

Protected areas in North America data (2021), downloaded from:
<http://www.cec.org/north-american-environmental-atlas/north-american-protected-areas-2021/>

##### IUCN categories:

-   Ia. Strict Nature Reserve (managed mainly for science or wilderness
    protection)
-   Ib. Wilderness Area (managed mainly for science or wilderness
    protection)
-   II. National Park (managed mainly for ecosystem protection and
        recreation)
-   III. Natural Monument or Feature (managed mainly for conservation of
         specific natural features)
-   IV. Habitat/Species Management Area (managed mainly for conservation
        through management intervention)
-   V. Protected Landscape/Seascape (managed mainly for
    landscape/seascape protection and recreation)
-   VI. Protected area with sustainable use of natural resources
        (managed mainly for sustainable use of natural systems)

```{r, echo = FALSE}
protect_area_IUCN_sf <- read_sf("data/raw/CEC_NA_2021_terrestrial_IUCN_categories.shp")
protect_area_IUCN_vect <- vect(protect_area_IUCN_sf)
```

OECMs - Other Effective area-based Conservation Measures (not
categorized by IUCN), also downloaded from
<http://www.cec.org/north-american-environmental-atlas/north-american-protected-areas-2021/>

```{r, echo = FALSE}
protect_area_OECM_sf <- read_sf("data/raw/CEC_NA_2021_terrestrial_OECMs.shp")
protect_area_OECM_vect <- vect(protect_area_OECM_sf)
```

North American Climate Zones data (2021), downloaded from:
<http://www.cec.org/north-american-environmental-atlas/climate-zones-of-north-america/>
Issues downloading: apparently the file doesn't exist? Even though it's in the directory. Troubleshooting this
```{r, echo = FALSE}
climate_zones_sf <- read_sf("data/raw/North_America_Climate_Zones.shp")
climate_zones_vect <- vect(climate_zones_sf)
```

The below code is in progress, need to figure out which model to use,
the 30 degree lon/lat tile to use (or download global data), as well as
which SSPs are associated with the 126, 245, 370, 545 codes. These codes
seem to only exist within R and are not used in the CMIP6 reports. I put
my best guess in brackets following the SSP code for which scenario it
is, but haven't yet confirmed this

#### **Future**

Shared Socioeconomic Pathway 126 (no climate policy?)

SSP 126 for years 2021-2040:

```{r, echo = FALSE}
cmip6_2021_2040_126 <- geodata::cmip6_tile(lon, lat, 
                                           model, 
                                           ssp = "126", 
                                           time = "2021-2040", 
                                           var = , 
                                           res = 2.5, 
                                           path = "data/raw/")
```

SSP 126 for years 2041-2060:

```{r, echo = FALSE}
cmip6_2041_2060_126 <- geodata::cmip6_tile(lon, lat, 
                                           model, 
                                           ssp = "126", 
                                           time = "2041-2060", 
                                           var = , 
                                           res = 2.5, 
                                           path = "data/raw/")
```

SSP 126 for years 2061-2080:

```{r, echo = FALSE}
cmip6_2061_2080_126 <- geodata::cmip6_tile(lon, lat, 
                                           model, 
                                           ssp = "126", 
                                           time = "2061-2080", 
                                           var = , 
                                           res = 2.5, 
                                           path = "data/raw/")
```

Future climate predictions shared Socioeconomic Pathway 245 (low/mid?)

SSP 245 for years 2021-2040:

```{r, echo = FALSE}
cmip6_2021_2040_245 <- geodata::cmip6_tile(lon, lat, 
                                           model, 
                                           ssp = "245",
                                           time = "2021-2040",
                                           var = , 
                                           res = 2.5, 
                                           path = "data/raw/")
```

SSP 245 for years 2041-2060:

```{r, echo = FALSE}
cmip6_2041_2060_245 <- geodata::cmip6_tile(lon, lat,
                                           model, 
                                           ssp = "245", 
                                           time = "2041-2060",
                                           var = , 
                                           res = 2.5, 
                                           path = "data/raw/")
```

SSP 245 for years 2061-2080:

```{r, echo = FALSE}
cmip6_2061_2080_245 <- geodata::cmip6_tile(lon, lat, 
                                           model, 
                                           ssp = "245", 
                                           time = "2061-2080", 
                                           var = , 
                                           res = 2.5, 
                                           path = "data/raw/")
```

Future climate predictions Shared Socioeconomic Pathway 370 (mid/high?)

SSP 370 for years 2021-2040:

```{r, echo = FALSE}
cmip6_2021_2040_370 <- geodata::cmip6_tile(lon, lat, 
                                           model, 
                                           ssp = "370",
                                           time = "2021-2040",
                                           var = ,
                                           res = 2.5,
                                           path = "data/raw/")
```

SSP 370 for years 2041-2060"

```{r, echo = FALSE}
cmip6_2041_2060_370 <- geodata::cmip6_tile(lon, lat, 
                                           model, 
                                           ssp = "370", 
                                           time = "2041-2060",
                                           var = , 
                                           res = 2.5, 
                                           path = "data/raw/")
```

SSP 370 for years 2061-2080:

```{r, echo = FALSE}
cmip6_2061_2080_370 <- geodata::cmip6_tile(lon, lat, 
                                           model, 
                                           ssp = "370", 
                                           time = "2061-2080",
                                           var = , 
                                           res = 2.5, 
                                           path = "data/raw/")
```

Future climate predictions Shared Socioeconomic Pathway 545 (high?)

SSP 545 for years 2021-2040:

```{r, echo = FALSE}
cmip6_2021_2040_545 <- geodata::cmip6_tile(lon, lat,
                                           model, 
                                           ssp = "545", 
                                           time = "2021-2040",
                                           var = , 
                                           res = 2.5, 
                                           path = "data/raw/")
```

SSP 545 for years 2041-2060:

```{r, echo = FALSE}
cmip6_2041_2060_545 <- geodata::cmip6_tile(lon, lat,
                                           model, 
                                           ssp = "545",
                                           time = "2041-2060",
                                           var = , 
                                           res = 2.5,
                                           path = "data/raw/")
```

SSP 545 for years 2061-2080: \`\`\`{r, echo = FALSE} cmip6_2061_2080_545
\<- geodata::cmip6_tile(lon, lat, model, ssp = "545", time =
"2061-2080", var = , res = 2.5, path = "data/raw/")
